---
title: "data_HuukiMyers"
format: html
editor: visual
---

# 0. Environment Set-Up

Packages
```{r}
#| label: libraries
#| output: false
#| warning: false
library(SingleCellExperiment)
library(ggplot2)
library(httr)
library(data.table)

```

```{r}
#| label: global_variables
data_dir <- "../data/"
dir.exists(data_dir)

source("../R/Omega_tensor.R")

RECOMPUTE_PREPROCESSING <- FALSE

ntfy_topic <- "A01422880"
```


# 1. Single-Cell Data Pre-Processing

Loading full snRNA-seq dataset
```{r}
#| label: load_full_sce
#| eval: !expr RECOMPUTE_PREPROCESSING
#| output: false
#| warning: false

if (RECOMPUTE_PREPROCESSING) {
  # Checking if Huuki-Myers data directory exists
  if(!dir.exists(paste0(data_dir, "Data_Huuki-Myers2025_Brain/"))){
    sce_path_zip <- DeconvoBuddies::fetch_deconvo_data("sce")
    unzip(sce_path_zip, exdir = paste0(data_dir, "Data_Huuki-Myers2025_Brain/"))
    rm(sce_path_zip)
  }
  
  # Loading SingleCellExperiment object
  sce <- HDF5Array::loadHDF5SummarizedExperiment(file.path(paste0(data_dir, "Data_Huuki-Myers2025_Brain/"), "sce_DLPFC_annotated"))
}
```

Getting lists of all cell types and cell states and removing cells tagges as "Ambiguous".
```{r}
#| label: pp_cell_lists
#| eval: !expr RECOMPUTE_PREPROCESSING

if (RECOMPUTE_PREPROCESSING) {
  # Fetching cell types
  all_types <- names(table(sce$cellType_broad_hc))
  
  # Removing ambiguous cell type
  all_types <- all_types[!all_types %in% c("Ambiguous")]
  sce <- sce[, sce$cellType_broad_hc %in% all_types]
  
  # Fetching cell states
  all_states <- unique(as.vector(sce$cellType_hc))
}
```

Graphing correlation across cell states
```{r}
#| label: fig-state_correlations
#| eval: !expr RECOMPUTE_PREPROCESSING
#| fig-cap: "Heatmap of expression correlation across cell states."
#| cache: true

if (RECOMPUTE_PREPROCESSING) {
BayesPrism::plot.cor.phi (input=t(counts(sce)),
                input.labels=sce$cellType_hc,
                title="cell state correlation",
                #specify pdf.prefix if need to output to pdf
                pdf.prefix="gbm.cor.cs", 
                cexRow=0.2, cexCol=0.2,
                margins=c(2,2))
}
```

Graphing correlation across cell types
```{r}
#| label: fig-type_correlations
#| eval: !expr RECOMPUTE_PREPROCESSING
#| fig-cap: "Heatmap of expression correlation across cell types."
#| cache: true
if (RECOMPUTE_PREPROCESSING){
  BayesPrism::plot.cor.phi (input=t(counts(sce)),
                input.labels=sce$cellType_broad_hc, 
                title="cell type correlation",
                #specify pdf.prefix if need to output to pdf
                #pdf.prefix="gbm.cor.ct",
                cexRow=0.5, cexCol=0.5,
                )
}
```

Getting and plotting gene statistics
```{r}
#| label: fig-specificty
#| eval: !expr RECOMPUTE_PREPROCESSING
#| fig-cap: "Gene specificity plot. Ribosomal, mitochondrial and chrX  chrY genes have high abundance but low specificity. They may be excluded from the analysis."
#| cache: true

if (RECOMPUTE_PREPROCESSING) {
  rownames(sce) <- make.unique(rownames(sce))
  sc.stat <- BayesPrism::plot.scRNA.outlier(
    input=t(assay(sce)), #make sure the colnames are gene symbol or ENSMEBL ID 
    cell.type.labels=sce$cellType_broad_hc,
    species="hs", #currently only human(hs) and mouse(mm) annotations are supported
    return.raw=TRUE #return the data used for plotting. 
    #pdf.prefix="gbm.sc.stat" specify pdf.prefix if need to output to pdf
  )
  rowData(sce) <- cbind(rowData(sce), sc.stat)
  rm(sc.stat)
}
```

Gene-level QC
```{r}
#| label: gene_qc
#| eval: !expr RECOMPUTE_PREPROCESSING

if (RECOMPUTE_PREPROCESSING) {
  # Removing ribosomal, mitochondrial and chrX  chrY genes
  sc.dat.filtered <- BayesPrism::cleanup.genes(input=t(assay(sce)),
                                               input.type="count.matrix",
                                               species="hs", 
                                               gene.group=c( "Rb","Mrp","other_Rb","chrM","MALAT1","chrX","chrY") ,
                                               exp.cells=5)
  sce <- sce[colnames(sc.dat.filtered), ]
  rm(sc.dat.filtered)
  
  # Limiting to protein coding genes
  sc.dat.filtered.pc <-  BayesPrism::select.gene.type(t(assay(sce)),
                                                      gene.type = "protein_coding")
  sce <- sce[colnames(sc.dat.filtered.pc), ]
  rm(sc.dat.filtered.pc)
  
  # Save the object
  HDF5Array::saveHDF5SummarizedExperiment(
    x = sce, 
    dir = paste0(data_dir, "NotebookRData/data_HuukiMyers/pp_sce"), 
    replace = TRUE,     # Overwrite if it already exists
    chunkdim = NULL,    # Let the package decide optimal chunking
    level = NULL        # Default compression level
  )
}
```

## 1.1 Loading Pre-Processed Single-Cell Data
```{r}
#| label: load_preprocessed
#| eval: !expr "!RECOMPUTE_PREPROCESSING"
sce <- HDF5Array::loadHDF5SummarizedExperiment(
  dir = paste0(data_dir, "NotebookRData/data_HuukiMyers/pp_sce")
)
```

```{r}
#| label: cell_lists

# Standardizing metadata for generalizability
sce$deconv_type  <- sce$cellType_broad_hc
sce$deconv_state <- sce$cellType_hc
sce$ref_sample <- sce$Sample

# Drop empty factors
sce$deconv_state <- droplevels(sce$deconv_state)
sce$deconv_type <- droplevels(sce$deconv_type)

# Get lists of all types and all states
all_types <- unique(as.vector(sce$deconv_type))
all_states <- unique(as.vector(sce$deconv_state))

# Get type to state map
type_to_state <- lapply(all_types,function(t)unique(as.vector(sce$deconv_state[sce$deconv_type==t])))
names(type_to_state) <- all_types

# Get state to type map
state_to_type <- stack(type_to_state)
colnames(state_to_type) <- c("cell_state", "cell_type")
```

Color dictionaries for graphs
```{r}
#| label: color_dicts
type_colors <- c(
  Inhib = "#369acc", 
  Oligo = "#f8e16f", 
  OPC = "#6c584c", 
  Excit = "#95cf92", 
  Astro = "#9656a2", 
  EndoMural = "#f4895f", 
  Micro = "#de324c"
)

#type_colors <- setNames(brewer.pal(length(type_to_state), "Set3"), names(type_to_state))

state_colors <- unlist(lapply(names(type_to_state), function(ct) {
  states <- sort(type_to_state[[ct]])
  n <- length(states)
  base <- type_colors[ct]
  
  if (n == 1) return(setNames(base, states))
  
  pal_func <- colorRampPalette(c(
    colorspace::lighten(base, 0.4),
    base, 
    colorspace::darken(base, 0.4)
  ))
  
  # 3. Assign the gradient to the sorted names
  # This ensures Excit_01 gets the lightest and Excit_15 gets the darkest
  return(setNames(pal_func(n), states))
}))

#scales::show_col(type_colors)
#scales::show_col(state_colors)
```

```{r}
#| label: ref_data_tag
ref_data_tag <- "HMHM"
```


# 2. Bulk Data
```{r}
#| label: blk_load
#| output: false
#| warning: false
blk <- DeconvoBuddies::fetch_deconvo_data("rse_gene")

# Ensure Gene Symbols are used as the rownames in bulk dataset
rownames(blk) <- rowData(blk)$Symbol
blk$ref_sample <- blk$Sample
blk$blk_sample <- blk$SAMPLE_ID
```

```{r}
#| label: mix_data_tag

mix_data_tag <- "blkHM"
```


# 3. Dataset-Specific Objects
```{r}
#| label: functions
source("../R/Omega_tensor.R")
```

## 3.1 Counts Per State and Sample
```{r}
#| label: get_state_sample_counts

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "state_sample_counts.rds"))){
  state_sample_counts <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "state_sample_counts.rds"))
} else {
  # Uses httr
  tryCatch({
    # Gets state_sample_counts list
    state_sample_counts <- aggregate_stratified_counts(t(assay(sce, "counts")), sce$ref_sample, sce$deconv_state, include_pseudobulk = TRUE)
    
    # Stores state_sample_counts
    saveRDS(state_sample_counts, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "state_sample_counts.rds"))
    
    # Uses nfty to notify process completion
    httr::POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved state_sample_counts successfully ✅",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    # Uses nfty to notify process error
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting state_sample_counts failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}
```

```{r}
gt_s_counts <- array(NA, dim = c(length(unique(state_sample_counts$meta$Sample)), 
                                   length(unique(rownames(state_sample_counts$pseudoexp))),
                                   length(unique(state_sample_counts$meta$Label))),
                       dimnames = list(unique(state_sample_counts$meta$Sample),
                                       unique(rownames(state_sample_counts$pseudoexp)),
                                       unique(state_sample_counts$meta$Label)))

for (col_name in colnames(state_sample_counts$pseudoexp)) {
  parts <- unlist(strsplit(col_name, ":::"))
  n <- parts[1]
  cs <- parts[2]
  
  gt_s_counts[n, , cs] <- state_sample_counts$pseudoexp[, col_name]
}

gt_s_logCP10k <- apply(gt_s_counts, c(1,3), function(x) log1p((x / sum(x)) * 1e4))
```


```{r}
#| label: get_type_sample_counts

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "type_sample_counts.rds"))){
  type_sample_counts <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "type_sample_counts.rds"))
} else {
  tryCatch({
    type_sample_counts <- aggregate_stratified_counts(t(assay(sce, "counts")), sce$ref_sample, sce$cellType_broad_hc, include_pseudobulk = TRUE)
    saveRDS(type_sample_counts, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "type_sample_counts.rds"))
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved type_sample_counts successfully ✅",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting type_sample_counts failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}
```

```{r}
#| label: ground_truth_type
gt_t_counts <- array(NA, dim = c(length(unique(type_sample_counts$meta$Sample)), 
                                   length(unique(rownames(type_sample_counts$pseudoexp))),
                                   length(unique(type_sample_counts$meta$Label))),
                       dimnames = list(unique(type_sample_counts$meta$Sample),
                                       unique(rownames(type_sample_counts$pseudoexp)),
                                       unique(type_sample_counts$meta$Label)))

for (col_name in colnames(type_sample_counts$pseudoexp)) {
  # Parse the sample and state for the current column
  parts <- unlist(strsplit(col_name, ":::"))
  n <- parts[1]
  ct <- parts[2]
  
  # Fill the slice: tensor[sample, all genes, cell state]
  gt_t_counts[n, , ct] <- type_sample_counts$pseudoexp[, col_name]
}

gt_t_logCP10k <- apply(gt_t_counts, c(1,3), function(x) log1p((x / sum(x)) * 1e4))
```

## 3.2 BayesPrism Results
### 3.2.1 New Prism

```{r}
#| label: new_prism

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "myPrism.rds"))){
  myPrism <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "myPrism.rds"))
} else {
  tryCatch({
    # Gets prism object
    myPrism <- BayesPrism::new.prism(
      reference=t(as.matrix(counts(sce))), 
      mixture=t(assay(blk)),
      input.type="count.matrix", 
      cell.type.labels = sce$deconv_type, 
      cell.state.labels = sce$deconv_state,
      key = NULL,
      outlier.cut=0.01,
      outlier.fraction=0.1,
      )
    
    # Stores prism object
    saveRDS(myPrism, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "myPrism.rds"))
    
    # Uses nfty to notify process completion
    httr::POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved myPrism successfully ✅",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    # Uses nfty to notify process error
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting myPrism failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

### 3.2.2 Run BayesPrism

```{r}
#| label: bayesprism

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "bp_blk.rds"))){
  bp_blk <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "bp_blk.rds"))
} else {
  tryCatch({
    # Run BayesPrism on sce
    bp_blk <- BayesPrism::run.prism(prism = myPrism, n.cores=10)
    
    # Stores BayesPrism results
    saveRDS(bp_blk, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "bp_blk.rds"))
    
    # Uses nfty to notify process completion
    httr::POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Ran BayesPrim successfully ✅",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    # Uses nfty to notify process error
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Running BayesPrism failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

## 3.3 Correlations

```{r}
#| eval: false

build_C_slg <- function(tCorr, rho_threshold = 0,
                         ligand_filter = NULL, gene_filter = NULL) {
  
  # Determine dimensions — pre-filter to relevant ligands/genes if provided
  # to keep the tensor as small as possible
  all_receivers <- unique(tCorr$Receiver)
  all_ligands   <- unique(tCorr$Ligand)
  all_genes     <- unique(tCorr$Gene)
  
  if (!is.null(ligand_filter)) all_ligands <- intersect(all_ligands, ligand_filter)
  if (!is.null(gene_filter))   all_genes   <- intersect(all_genes,   gene_filter)
  
  message(sprintf("Allocating C_slg: %d receivers x %d ligands x %d genes (%.1f MB)",
                  length(all_receivers), length(all_ligands), length(all_genes),
                  length(all_receivers) * length(all_ligands) * length(all_genes) * 8 / 1e6))
  
  # Pre-allocate tensor [Receiver x Ligand x Gene]
  C_slg <- array(0,
                 dim = c(length(all_receivers), length(all_ligands), length(all_genes)),
                 dimnames = list(Receiver = all_receivers,
                                 Ligand   = all_ligands,
                                 Gene     = all_genes))
  
  # Process one receiver at a time to avoid copying full tCorr
  for (rec in all_receivers) {
    message(sprintf("Processing %s...", rec))
    
    rec_dt <- tCorr[Receiver == rec & 
                    !is.na(Rho)          & 
                    abs(Rho) >= rho_threshold &
                    Ligand %in% all_ligands   &
                    Gene   %in% all_genes]
    
    if (nrow(rec_dt) == 0) {
      message(sprintf("  No correlations above threshold for %s, skipping.", rec))
      next
    }
    
    # Cast to [Ligand x Gene] matrix using raw Rho
    rec_wide <- dcast(rec_dt, Ligand ~ Gene, value.var = "Rho", fill = 0)
    
    ligs <- rec_wide$Ligand
    m    <- as.matrix(rec_wide[, !"Ligand"])
    rownames(m) <- ligs
    
    # Fill tensor slice
    common_ligs  <- intersect(all_ligands, rownames(m))
    common_genes <- intersect(all_genes,   colnames(m))
    
    C_slg[rec, common_ligs, common_genes] <- m[common_ligs, common_genes]
    
    rm(rec_dt, rec_wide, m, ligs)
    gc()
  }
  
  return(C_slg)
}
```

### 3.3.1 State-Level

```{r}
#| label: state_corrs
#| eval: false

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tCorr_s.rds"))){
  tCorr_s <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tCorr_s.rds"))
} else {
  tryCatch({
    tCorr_s <- get_trans_correlations_parallel(expr_mat = state_sample_counts$logexpr, 
                                             meta = state_sample_counts$meta, 
                                             lig_genes = lig_genes, 
                                             tar_genes = tar_genes, 
                                             n_cores = 8, 
                                             rho_threshold = 0)
    saveRDS(tCorr_s, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tCorr_s.rds"))
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved corr_s successfully",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting corr_s failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

```{r}
#| label: C_slg
#| eval: false

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "C_slg.rds"))){
  C_slg <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "C_slg.rds"))
} else {
  tryCatch({
    C_slg <- get_trans_correlations_parallel(expr_mat = state_sample_counts$logexpr, 
                                             meta = state_sample_counts$meta, 
                                             lig_genes = lig_genes, 
                                             tar_genes = tar_genes, 
                                             n_cores = 8, 
                                             rho_threshold = 0)
    saveRDS(C_slg, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "C_slg.rds"))
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved C_slg successfully",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting C_slg failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

```{r}
#| label: state_consensus

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tConsensus_s.rds"))){
  tConsensus_s <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tConsensus_s.rds"))
} else {
  tryCatch({
    tConsensus_s <- compute_stouffer_consensus(tCorr_s)
    saveRDS(tConsensus_s, file = paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tConsensus_s.rds"))
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved tConsensus_s successfully",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting tConsensus_s  failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

```{r}
#| label: state_C_lg
norm_factor <- quantile(abs(tConsensus_s$StoufferZ), 0.95, na.rm = TRUE)
tConsensus_s[, Z_norm := StoufferZ / norm_factor]

tConsensus_s$Z_filtered <- tConsensus_s$Z_norm
tConsensus_s[Consistency < 0.85, Z_filtered := 0]

C_s_tmp <- dcast(tConsensus_s, Ligand ~ Gene, value.var = "Z_filtered", fill = 0)

C_s_mtx <- as.matrix(C_s_tmp[, -1])
rownames(C_s_mtx) <- C_s_tmp$Ligand

rm(C_s_tmp)
```

### 3.3.1 Type-Level

```{r}
#| label: type_corrs
#| eval: false

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tCorr_t.rds"))){
  tCorr_t <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tCorr_t.rds"))
} else {
  tryCatch({
    tCorr_t <- get_trans_correlations_parallel(expr_mat = type_sample_counts$logexpr, 
                                             meta = type_sample_counts$meta, 
                                             lig_genes = lig_genes, 
                                             tar_genes = tar_genes, 
                                             n_cores = 8, 
                                             rho_threshold = 0)
    saveRDS(tCorr_t, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tCorr_t.rds"))
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved tCorr_t successfully",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting tCorr_t failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

```{r}
#| label: C_tlg
#| eval: false

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "C_tlg.rds"))){
  C_tlg <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "C_tlg.rds"))
} else {
  C_tlg <- build_C_slg(tCorr_t,
                        rho_threshold  = 0)
  
  saveRDS(C_tlg, file = paste0(data_dir, "NotebookRData/data_HuukiMyers/", "C_tlg.rds"))
}

```

```{r}
#| label: type_consensus

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tConsensus_t.rds"))){
  tConsensus_t <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tConsensus_t.rds"))
} else {
  tryCatch({
    tConsensus_t <- compute_stouffer_consensus(tCorr_t)
    saveRDS(tConsensus_t, file = paste0(data_dir, "NotebookRData/data_HuukiMyers/", "tConsensus_t.rds"))
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved tConsensus_t successfully",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting tConsensus_t  failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

```{r}
#| label: type_C_lg
norm_factor <- quantile(abs(tConsensus_t$StoufferZ), 0.95, na.rm = TRUE)
tConsensus_t[, Z_norm := StoufferZ / norm_factor]

tConsensus_t$Z_filtered <- tConsensus_t$Z_norm
tConsensus_t[Consistency < 0.85, Z_filtered := 0]

C_t_tmp <- dcast(tConsensus_t, Ligand ~ Gene, value.var = "Z_filtered", fill = 0)

C_t_mtx <- as.matrix(C_t_tmp[, -1])
rownames(C_t_mtx) <- C_t_tmp$Ligand

rm(C_t_tmp)
```


# 4. Pseudobulk

```{r}
psb_counts <- apply(gt_s_counts, c(1,2), sum, na.rm = TRUE)
```

```{r}
#| label: new_prism

if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "psbPrism.rds"))){
  psbPrism <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "psbPrism.rds"))
} else {
  tryCatch({
    # Gets prism object
    psbPrism <- BayesPrism::new.prism(
      reference=t(as.matrix(counts(sce))), 
      mixture=psb_counts,
      input.type="count.matrix", 
      cell.type.labels = sce$deconv_type, 
      cell.state.labels = sce$deconv_state,
      key = NULL,
      outlier.cut=0.01,
      outlier.fraction=0.1,
      )
    
    # Stores prism object
    saveRDS(psbPrism, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "psbPrism.rds"))
    
    # Uses nfty to notify process completion
    httr::POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Saved psbPrism successfully ✅",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    # Uses nfty to notify process error
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Getting psbPrism failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```

```{r}
#| label: pseudobulk_BayesPrism
#| 
if (file.exists(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "bp_psb.rds"))){
  bp_psb <- readRDS(paste0(data_dir, "NotebookRData/data_HuukiMyers/", "bp_psb.rds"))
} else {
  tryCatch({
    # Run BayesPrism on sce
    bp_psb <- BayesPrism::run.prism(prism = psbPrism, n.cores=10)
    
    # Stores BayesPrism results
    saveRDS(bp_psb, file=paste0(data_dir, "NotebookRData/data_HuukiMyers/", "bp_psb.rds"))
    
    # Uses nfty to notify process completion
    httr::POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = "Ran BayesPrim successfully ✅",
         add_headers(Title = "✅ R Process Success")
         )
  }, error = function(e) {
    # Uses nfty to notify process error
    POST(paste0("https://ntfy.sh/", ntfy_topic),
         body = paste0("Running BayesPrism failed: ", e$message),
         add_headers(Title = "❌ R Process ERROR", Priotity = "high")
         )
  })
}

```



