---
title: "HuukiResiduals"
format: html
editor: visual
---

# 0. Environment Set-Up

Packages

```{r}
#| label: libraries
#| output: false
#| warning: false
library(BayesPrism)
library(SingleCellExperiment)
library(biomaRt)
library(ggplot2)
library(ggalign)
library(tidyverse)
library(scuttle)
library(scales)
library(colorspace)
library(Hmisc)
library(data.table)
```

Functions

```{r}
#| label: functions
source("../R/Omega_tensor.R")
```

Global variables

```{r}
#| label: global_variables
data_dir <- "../data/"
dir.exists(data_dir)
```
```{r}
#| label: ntfy_setup
library("httr")
ntfy_topic <- "A01422880"
```


# 1. Loading Data

Reference data

```{r}
#| label: sce_load
#| output: false
#| warning: false
if(!dir.exists(paste0(data_dir, "Data_Huuki-Myers2025_Brain/"))){
  sce_path_zip <- DeconvoBuddies::fetch_deconvo_data("sce")
  unzip(sce_path_zip, exdir = paste0(data_dir, "Data_Huuki-Myers2025_Brain/"))
  rm(sce_path_zip)
}

sce <- HDF5Array::loadHDF5SummarizedExperiment(file.path(paste0(data_dir, "Data_Huuki-Myers2025_Brain/"), "sce_DLPFC_annotated"))
```

Matched bulk data

```{r}
#| label: blk_load
#| output: false
#| warning: false
blk <- DeconvoBuddies::fetch_deconvo_data("rse_gene")
```

NicheNet data

```{r}
#| label: NicheNet_load
# Ligand target matrix. Target genes in rows, ligands in columns.
ligand_target_matrix <- readRDS(paste0(data_dir, "nichenetr.dat/ligand_target_matrix_nsga2r_final.rds")) 
lr_network <- readRDS(paste0(data_dir, "nichenetr.dat/lr_network_human_21122021.rds"))
```

# 2. Reference quality control

Getting lists of cell types and cell states.

```{r}
#| label: cell_lists
# Removing ambiguous cell type
all_types <- names(table(sce$cellType_broad_hc))
all_types <- all_types[!all_types %in% c("Ambiguous")]
sce <- sce[, sce$cellType_broad_hc %in% all_types]
all_states <- unique(as.vector(sce$cellType_hc))
```

Graphing correlation across cell states

```{r}
#| label: fig-state_correlations
#| fig-cap: "Heatmap of expression correlation across cell states."
#| cache: true
plot.cor.phi (input=t(counts(sce)),
              input.labels=sce$cellType_hc,
              title="cell state correlation",
              #specify pdf.prefix if need to output to pdf
              #pdf.prefix="gbm.cor.cs", 
              cexRow=0.2, cexCol=0.2,
              margins=c(2,2))
```

Graphing correlation across cell types

```{r}
#| label: fig-type_correlations
#| fig-cap: "Heatmap of expression correlation across cell types."
#| cache: true
plot.cor.phi (input=t(counts(sce)),
              input.labels=sce$cellType_broad_hc, 
              title="cell type correlation",
              #specify pdf.prefix if need to output to pdf
              #pdf.prefix="gbm.cor.ct",
              cexRow=0.5, cexCol=0.5,
              )
```

Getting and plotting gene statistics

```{r}
#| label: fig-specificty
#| fig-cap: "Gene specificity plot. Ribosomal, mitochondrial and chrX  chrY genes have high abundance but low specificity. They may be excluded from the analysis."
#| cache: true
rownames(sce) <- make.unique(rownames(sce))
sc.stat <- plot.scRNA.outlier(
  input=t(assay(sce)), #make sure the colnames are gene symbol or ENSMEBL ID 
  cell.type.labels=sce$cellType_broad_hc,
  species="hs", #currently only human(hs) and mouse(mm) annotations are supported
  return.raw=TRUE #return the data used for plotting. 
  #pdf.prefix="gbm.sc.stat" specify pdf.prefix if need to output to pdf
)
rowData(sce) <- cbind(rowData(sce), sc.stat)
rm(sc.stat)
```

Gene-level QC

```{r}
#| label: gene_qc
#| eval: false
# Removing ribosomal, mitochondrial and chrX  chrY genes
sc.dat.filtered <- cleanup.genes (input=t(assay(sce)),
                                  input.type="count.matrix",
                                    species="hs", 
                                    gene.group=c( "Rb","Mrp","other_Rb","chrM","MALAT1","chrX","chrY") ,
                                    exp.cells=5)
sce <- sce[colnames(sc.dat.filtered), ]
rm(sc.dat.filtered)

# Limiting to protein coding genes
sc.dat.filtered.pc <-  select.gene.type (t(assay(sce)),
                                        gene.type = "protein_coding")
sce <- sce[colnames(sc.dat.filtered.pc), ]
rm(sc.dat.filtered.pc)
save(sce, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "sce.rdata"))
```

Loading QC-filtered SingeCellExperiment object (`sce`)

```{r}
#| label: load_qc
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", "sce.rdata"))
```

# 3. BayesPrism

Defining sample

```{r}
#| label: sample_vars
br_num = "Br2720"
position = "post"
sample = paste0(br_num,"_",position)

rownames(blk) <- rowData(blk)$Symbol
blk_sample = blk[intersect(rownames(sce), rownames(blk)),blk$Sample==sample]
```

Initializing Prism object

```{r}
#| label: new_prism
#| eval: false
myPrism <- new.prism(
  reference=t(as.matrix(counts(sce[rownames(blk_sample),]))), 
  mixture=t(assay(blk_sample)),
  input.type="count.matrix", 
  cell.type.labels = sce$cellType_broad_hc, 
  cell.state.labels = sce$cellType_hc,
  key = NULL,
  outlier.cut=0.01,
    outlier.fraction=0.1,
)
save(myPrism, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "myPrism_", sample, ".rdata"))
```

Running BayesPrism

```{r}
#| label: bayesprism
#| eval: false
bp_res <- run.prism(prism = myPrism, n.cores=14)
save(bp_res, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "bp_res_", sample, ".rdata"))
```

Loading stored BayesPrism prism object (`myPrism`) and results (`bp_res`)

```{r}
#| label: load_bayesprism
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", "myPrism_", sample, ".rdata"))
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", "bp_res_", sample, ".rdata"))
```

Color Mapping

```{r}
#| label: color_dicts
type_colors <- c(
  Inhib = "#369acc", 
  Oligo = "#f8e16f", 
  OPC = "#6c584c", 
  Excit = "#95cf92", 
  Astro = "#9656a2", 
  EndoMural = "#f4895f", 
  Micro = "#de324c"
)

state_colors <- unlist(lapply(names(bp_res@prism@map), function(cell_type) {
  states <- bp_res@prism@map[[cell_type]]
  base_col <- type_colors[cell_type]
  
  shades <- colorRampPalette(c(lighten(base_col, 0.4), darken(base_col, 0.4)))(length(states))
  
  names(shades) <- states
  return(shades)
}))
```

# 4. Calculating Residuals

Bulk sample selection

```{r}
#| label: lib_vars
lib_prep <- "Bulk"
lib_type <- "polyA"
```

Calculating expressions per cell state

-   `inf_exp_mtx`: BayesPrism-inferred CTSE in bulk RNA-seq data from `{r} sample`
-   `true_exp_mtx`: CTSE in matched snRNA-seq data from `{r} sample`

```{r}
#| label: exp_mtxs
#| eval: false
inf_exp_mtx <- matrix(nrow = length(all_states), 
                       ncol = length(colnames(bp_res@prism@mixture)),
                       dimnames = list(all_states, colnames(bp_res@prism@mixture)))

true_exp_mtx <- matrix(nrow = length(all_states), 
                       ncol = length(rownames(sce)),
                       dimnames = list(all_states, rownames(sce)))

for (rec in all_states){
  ids <- colnames(blk_sample[,blk_sample$library_prep == lib_prep & blk_sample$library_type == lib_type])
  rec_idx <- colnames(sce)[sce$cellType_hc == rec & sce$Sample == sample]
  
  inferred_rec_exp <- get.exp(bp_res, state.or.type = "state", cell.name = rec)[ids,colnames(inf_exp_mtx)]
  true_rec_exp <- rowSums(as.matrix(counts(sce[colnames(true_exp_mtx), rec_idx])))
  
  inf_exp_mtx[rec, ] <- inferred_rec_exp
  true_exp_mtx[rec, ] <- true_rec_exp
}

save(inf_exp_mtx, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", 
                              "inf_exp_mtx_", sample, "_", lib_prep, "_", lib_type,".rdata"))
save(true_exp_mtx, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", 
                               "true_exp_mtx_", sample, "_", lib_prep, "_", lib_type,".rdata"))
```

Loading stored expression matrices

```{r}
#| label: load_exp_mtxs
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", 
                              "inf_exp_mtx_", sample, "_", lib_prep, "_", lib_type,".rdata"))
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", 
                              "true_exp_mtx_", sample, "_", lib_prep, "_", lib_type,".rdata"))
```

Calculating CPM and residuals

```{r}
#| label: residual_calc
inf_cpm_mtx <- (inf_exp_mtx/rowSums(inf_exp_mtx)) * 1e4
true_cpm_mtx <- (true_exp_mtx/rowSums(true_exp_mtx)) * 1e4

residual_mtx <- true_cpm_mtx[all_states, colnames(inf_cpm_mtx)] - inf_cpm_mtx[all_states, ]
```

# 5. Getting Omega Tensor

## 5.1 Getting average expression matrices

Generating and saving average expression matrices

```{r}
#| label: avg_exp
#| eval: false
# Getting average expression per cell type
avg_type_expr <- avg_per_label(sce, sce$cellType_broad_hc, assay = "logcounts")
save(avg_type_expr, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "avg_type_expr.rdata"))

# Getting average expression per cell state
avg_state_expr <- avg_per_label(sce, sce$cellType_hc, assay = "logcounts")
save(avg_state_expr, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "avg_state_expr.rdata"))
```

Loading average expression matrices

```{r}
#| label: load_avg_exp
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", "avg_type_expr.rdata"))
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", "avg_state_expr.rdata"))
```

## 5.2a Directionality

Getting average expression per sample per cell type

```{r}
#| label: state_sample
#| eval: false

state_sample_means <- aggregateAcrossCells(sce, 
                                           ids = colData(sce)[, c("Sample", "cellType_hc")], 
                                           statistics = "mean",
                                           use.assay.type = "logcounts")
save(state_sample_means, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "state_sample_means.rdata"))
```

```{r}
#| label: load_state_sample
load(paste0(data_dir, "NotebookRData/HuukiResiduals/", "state_sample_means.rdata"))
```

```{r}
#| label: state_sample_mtx
expr_mat <- assays(state_sample_means)$logcounts
meta <- as.data.frame(colData(state_sample_means))
```

```{r}
#| label: load_type_sample_counts
type_sample_counts <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "type_sample_counts.rds"))
```



```{r}
# Calculate Spearman correlation for each ligand
cor_results <- cor_df %>%
  group_by(lig) %>%
  dplyr::summarize(
    spearman_rho = cor(lig_exp, tar_exp, method = "spearman", use = "complete.obs"),
    p_value = cor.test(lig_exp, tar_exp, method = "spearman")$p.value
  ) %>%
  arrange(desc(spearman_rho))

C_tsr[receiver, cor_results$lig, target] <- cor_results$spearman_rho
```

```{r}
cor_df <- merge(
  cor_df, 
  cor_results, 
  by = "lig"
)
```

```{r}
plot_df <- cor_df[cor_df$p_value < 0.002 & !is.na(cor_df$spearman_rho),]

ggplot(plot_df, aes(x = tar_exp, y = lig_exp, colour = lig)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm") +
  labs(
    x = paste0(target, " logcounts on ", receiver)
  )
```

Getting ligand-target correlations across cell states

```{r}
#| label: corr_func
ligand_genes <- intersect(colnames(ligand_target_matrix), colnames(avg_state_expr))
target_genes <- intersect(rownames(ligand_target_matrix), colnames(avg_state_expr))
valid_states <- names(table(state_sample_means$cellType_hc)[table(state_sample_means$cellType_hc) > 4])

sub_expr <- expr_mat[unique(c(ligand_genes, target_genes)), ]

# Function to get correlations for a specific cell state
get_state_correlations <- function(cell_state) {
  cells <- which(meta$cellType_hc == cell_state)
  
  mat <- t(as.matrix(sub_expr[, cells]))
  
  cor_obj <- rcorr(mat, type = "spearman")
  
  r_matrix <- cor_obj$r[ligand_genes, target_genes]
  p_matrix <- cor_obj$P[ligand_genes, target_genes]
  
  df <- data.frame(
    ligand = rep(rownames(r_matrix), ncol(r_matrix)),
    target = rep(colnames(r_matrix), each = nrow(r_matrix)),
    cor    = as.vector(r_matrix),
    pval   = as.vector(p_matrix),
    state  = cell_state,
    stringsAsFactors = FALSE
  )
  
  return(df)
}
```

```{r}
#| label: run_corr
for (i in seq_along(valid_states)) {
  current_state <- valid_states[i]
  
  cat(sprintf("[%d/%d] Processing State: %s... \n", i, length(valid_states), current_state))
  
  if (file.exists(paste0(data_dir, "NotebookRData/HuukiResiduals/correlation_results/", 
                            current_state, "_cors.rds"))){next}
  
  res <- get_state_correlations(current_state)
  
  saveRDS(res, file = paste0(data_dir, "NotebookRData/HuukiResiduals/correlation_results/", 
                             current_state, "_cors.rds"))
  
  rm(res)
  gc()
}
```

```{r}
#| label: load_corr
#| eval: false
result_files <- list.files(paste0(data_dir, "NotebookRData/HuukiResiduals/correlation_results/"), pattern = "\\.rds$", full.names = TRUE)

all_cors_dt <- rbindlist(lapply(result_files, function(f) {
  tmp <- readRDS(f)
  
  tmp <- tmp[!is.na(tmp$cor),]
  return(as.data.table(tmp)) 
}))

setkey(all_cors_dt, ligand, target, state)

x <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/correlation_results/", "Astro_cors.rds"))
```

Define directionality consensus scores across cell states

```{r}
#| label: get_dir_summ
#| eval: false

consensus_summary <- all_cors_dt[, {
  z_vals <- qnorm(pmax(pval, 1e-300) / 2, lower.tail = FALSE) * sign(cor)
  
  stouffer_z <- sum(z_vals, na.rm = TRUE) / sqrt(.N)
  
  pos_count <- sum(cor > 0, na.rm = TRUE)
  neg_count <- sum(cor < 0, na.rm = TRUE)
  consistency <- max(pos_count, neg_count) / .N
  
  global_dir <- ifelse(pos_count >= neg_count, 1, -1)
  
  .(
    n_states      = .N,
    consensus_z   = stouffer_z,
    consistency   = consistency,
    global_dir    = global_dir,
    avg_cor       = mean(cor, na.rm = TRUE)
  )
}, by = .(ligand, target)]

save(consensus_summary, file=paste0(data_dir, "NotebookRData/Huukiresiduals/", "consensus_summary.rdata"))

# Filter for Significance
# A Z-score > 1.96 or < -1.96 roughly corresponds to p < 0.05
#conconsensus_summary[, final_pval := 2 * pnorm(abs(consensus_z), lower.tail = FALSE)]
```

```{r}
#| label: load_dir_summ
load(paste0(data_dir, "NotebookRData/Huukiresiduals/", "consensus_summary.rdata"))
```

Visualizing consensus data
```{r}
stack_vertical() +
  ggplot(consensus_summary) +
  geom_density(mapping = aes(consensus_z/max(abs(consensus_summary$consensus_z), na.rm = TRUE))) +
  labs(x = "MaxAbs-Normalized Cis-Z") +
  
  ggplot(consensus_summary) +
  geom_density(mapping = aes(avg_cor)) +
  labs(x = "Average Cis-Correlation")
```

```{r}
set.seed(111) # For reproducibility

norm_constant <- quantile(abs(consensus_summary$consensus_z), 0.95, na.rm = TRUE)

# 1. Pick 10 random unique ligands
subset_ligands <- unique(consensus_summary$ligand)# c(names(sort(rowMeans(cC_mtx, na.rm = TRUE), decreasing = TRUE)[c(1:3)]), names(sort(abs(rowMeans(cC_mtx, na.rm = TRUE)), decreasing = FALSE)[c(1:3)]), names(sort(rowMeans(cC_mtx, na.rm = TRUE), decreasing = FALSE)[c(1:3)]))#sample(unique(consensus_summary$ligand), 1000) 
#

# 2. Pick 100 random unique targets
subset_targets <- sample(unique(consensus_summary$target), 2000) # c(names(sort((colMeans(cC_mtx, na.rm = FALSE)), decreasing = TRUE)[c(1:100)]), names(sort((colMeans(cC_mtx, na.rm = FALSE)), decreasing = FALSE)[c(1:100)]))
  # unique(consensus_summary$target)#

# 3. Filter the main table for these specific groups
subset_dt <- consensus_summary[ligand %in% subset_ligands & target %in% subset_targets
                               & consistency >= 0.5
                               & consensus_z <= quantile(abs(consensus_summary$consensus_z), 0.99, na.rm = TRUE)
                               ]

ggside(subset_dt, mapping = aes(x = consensus_z/norm_constant, y = avg_cor, colour = consistency, alpha = n_states, size = n_states)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_point() +
  scale_color_viridis_c(option = "turbo") +
  scale_size_continuous(range = c(0.2, 2.5)) + 
  scale_alpha_continuous(range = c(0.3, 0.8)) +
  #scale_color_gradient(low = "grey", high = "darkgreen") +
  labs(x = "Q0.95Abs-Normalized Cis-Z", y = "Average Cis-Correlation") +
  
  anno_right(size = 0.3) +
  ggalign() +
  geom_density(aes(x = after_stat(density), y = avg_cor), position = "stack")+
  
  anno_top(size = 0.3) +
  ggalign() +
  geom_density(aes(x = consensus_z/norm_constant, y = after_stat(density)), position = "stack")
```



Aggregating consensus scores into a directionality matrix

```{r}
#| label: dir_mtx
directionality_dt <- dcast(
  consensus_summary, 
  ligand ~ target, 
  value.var = "consensus_z",
  fill = NA 
)

directionality_mtx <- as.matrix(directionality_dt[, -1])
rownames(directionality_mtx) <- directionality_dt$ligand
directionality_mtx <- directionality_mtx[ligand_genes, target_genes]
```

Getting a sign matrix

```{r}
#| label: sign_mtx
sign_mtx <- sign(directionality_mtx)
sign_mtx[which(is.na(sign_mtx))] <- 1
sign_mtx[which(sign_mtx == 0)] <- 1

W_matrix <- t(ligand_target_matrix[target_genes, ligand_genes])
signed_W_mtx <- sign_mtx * W_matrix # Provide this instead of ligand_target_matrix in build_Omega()
```

Modifying W matrix with consensus scores
```{r}
#| label: norm_dir
base_dir <- quantile(abs(directionality_mtx), 0.95, na.rm = TRUE)
norm_dir_mtx <- directionality_mtx/base_dir

norm_dir_mtx[which(is.na(norm_dir_mtx))] <- mean(norm_dir_mtx, na.rm=TRUE)

cor_W_mtx <- norm_dir_mtx * W_matrix
```

## 5.2b Context Matrix

### 5.2b.1 State-level

```{r}
#| label: get_state_sample_counts
tryCatch({
  state_sample_counts <- aggregate_stratified_counts(t(assay(sce, "counts")), sce$Sample, sce$cellType_hc, include_pseudobulk = TRUE)
  saveRDS(state_sample_counts, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "state_sample_counts.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved state_sample_counts successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting state_sample_counts failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```

```{r}
#| label: load_state_sample_counts
state_sample_counts <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "state_sample_counts.rds"))
```

```{r}
#| label: identify_ligs_n_tars
lig_genes <- intersect(colnames(ligand_target_matrix), rownames(state_sample_counts$logexpr))
tar_genes <- intersect(rownames(ligand_target_matrix), rownames(state_sample_counts$logexpr))
```

```{r}
#| label: get_state_cors
tryCatch({
  corr_s <- get_trans_correlations_parallel(expr_mat = state_sample_counts$logexpr, 
                                           meta = state_sample_counts$meta, 
                                           lig_genes = lig_genes, 
                                           tar_genes = tar_genes, 
                                           n_cores = 8, 
                                           rho_threshold = 0)
  saveRDS(corr_s, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "corr_s_dt.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved corr_s successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting corr_s failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```
```{r}
#| label: load_state_cors

corr_s <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "corr_s_dt.rds"))
```

```{r}
#| label: get_state_consensus
tryCatch({
  tConsensus_s <- compute_stouffer_consensus(corr_s)
  saveRDS(tConsensus_s, file = paste0(data_dir, "NotebookRData/HuukiResiduals/", "tConsensus_s.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved tConsensus_s successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting tConsensus_s  failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```

```{r}
#| label: load_state_consensus
tConsensus_s <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "tConsensus_s.rds"))
```

```{r}
#| label: visualize_state_consensus

set.seed(111) # For reproducibility

norm_constant <- quantile(abs(tConsensus_s$StoufferZ), 0.95, na.rm = TRUE)

subset_ligands <- unique(tConsensus_s$Ligand)# c(names(sort(rowMeans(cC_mtx, na.rm = TRUE), decreasing = TRUE)[c(1:3)]), names(sort(abs(rowMeans(cC_mtx, na.rm = TRUE)), decreasing = FALSE)[c(1:3)]), names(sort(rowMeans(cC_mtx, na.rm = TRUE), decreasing = FALSE)[c(1:3)]))#sample(unique(consensus_summary$ligand), 1000) 
#

subset_targets <- sample(unique(tConsensus_s$Gene), 2000) # c(names(sort((colMeans(cC_mtx, na.rm = FALSE)), decreasing = TRUE)[c(1:100)]), names(sort((colMeans(cC_mtx, na.rm = FALSE)), decreasing = FALSE)[c(1:100)]))
  # unique(consensus_summary$target)#

subset_dt <- tConsensus_s[Ligand %in% subset_ligands & Gene %in% subset_targets
                         #& StoufferZ <= quantile((tConsensus_s$StoufferZ), 0.95, na.rm = TRUE)
                         #& StoufferZ >= quantile((tConsensus_s$StoufferZ), 0.05, na.rm = TRUE)
                         ]

subset_dt <- subset_dt %>%
  mutate(Consistency_Range = cut(Consistency, 
                                 breaks = seq(0.5, 1.0, by = 0.5/7), 
                                 include.lowest = TRUE))
consistency_bins <- length(unique(subset_dt$Consistency_Range))

ggside(subset_dt, mapping = aes(x = StoufferZ/norm_constant, y = AvgRho, colour = Consistency, alpha = n_labels, size = n_labels)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_point() +
  scale_color_viridis_c(option = "turbo") +
  scale_size_continuous(range = c(0.2, 2.5)) + 
  scale_alpha_continuous(range = c(0.3, 0.8)) +
  #scale_color_gradient(low = "grey", high = "darkgreen") +
  labs(x = "Q0.95-Normalized Trans-Z", y = "Average Trans-Correlation") +
  
  anno_right(size = 0.3) +
  ggalign() +
  geom_density(aes(x = after_stat(density), y = AvgRho, 
                   colour = Consistency_Range), 
               position = "identity") +
  scale_color_manual(values = turbo(consistency_bins), guide = "none") +
  
  anno_top(size = 0.3) +
  ggalign() +
  geom_density(aes(x = StoufferZ/norm_constant, y = after_stat(density), 
                   colour = Consistency_Range), 
               position = "identity") +
  scale_color_manual(values = turbo(consistency_bins), guide = "none")
```

```{r}
#| label: get_state_C
norm_factor <- quantile(abs(tConsensus_s$StoufferZ), 0.95, na.rm = TRUE)
tConsensus_s[, Z_norm := StoufferZ / norm_factor]

C_s_tmp <- dcast(tConsensus_s, Ligand ~ Gene, value.var = "Z_norm", fill = 0)

C_s_mtx <- as.matrix(C_s_tmp[, -1])
rownames(C_s_mtx) <- C_s_tmp$Ligand

rm(C_s_tmp)
```

### 5.2b.2 Type-level

```{r}
#| label: get_type_sample_counts

tryCatch({
  type_sample_counts <- aggregate_stratified_counts(t(assay(sce, "counts")), sce$Sample, sce$cellType_broad_hc, include_pseudobulk = TRUE)
  saveRDS(type_sample_counts, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "type_sample_counts.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved type_sample_counts successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting type_sample_counts failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})

```

```{r}
#| label: load_type_sample_counts
type_sample_counts <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "type_sample_counts.rds"))
```

```{r}
#| label: get_type_cors
tryCatch({
  corr_t <- get_trans_correlations_parallel(expr_mat = type_sample_counts$logexpr, 
                                           meta = type_sample_counts$meta, 
                                           lig_genes = lig_genes, 
                                           tar_genes = tar_genes, 
                                           n_cores = 8, 
                                           rho_threshold = 0)
  saveRDS(corr_t, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "corr_t_dt.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved corr_t successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting corr_t failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```

```{r}
#| label: load_type_cors
corr_t <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "corr_t_dt.rds"))
```

```{r}
#| label: get_type_consensus
tryCatch({
  tConsensus_t <- compute_stouffer_consensus(corr_t)
  saveRDS(tConsensus_t, file = paste0(data_dir, "NotebookRData/HuukiResiduals/", "tConsensus_t.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved tConsensus_t successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting tConsensus_t  failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```

```{r}
#| label: load_type_consensus
tConsensus_t <- readRDS(paste0(data_dir, "NotebookRData/HuukiResiduals/", "tConsensus_t.rds"))
```

```{r}
#| label: visualize_type_consensus

set.seed(111) # For reproducibility

norm_constant <- quantile(abs(tConsensus_t$StoufferZ), 0.95, na.rm = TRUE)

subset_ligands <- unique(tConsensus_t$Ligand)# c(names(sort(rowMeans(cC_mtx, na.rm = TRUE), decreasing = TRUE)[c(1:3)]), names(sort(abs(rowMeans(cC_mtx, na.rm = TRUE)), decreasing = FALSE)[c(1:3)]), names(sort(rowMeans(cC_mtx, na.rm = TRUE), decreasing = FALSE)[c(1:3)]))#sample(unique(consensus_summary$ligand), 1000) 
#

subset_targets <- sample(unique(tConsensus_t$Gene), 2000) # c(names(sort((colMeans(cC_mtx, na.rm = FALSE)), decreasing = TRUE)[c(1:100)]), names(sort((colMeans(cC_mtx, na.rm = FALSE)), decreasing = FALSE)[c(1:100)]))
  # unique(consensus_summary$target)#

subset_dt <- tConsensus_t[Ligand %in% subset_ligands & Gene %in% subset_targets
                         #& StoufferZ <= quantile((tConsensus_s$StoufferZ), 0.95, na.rm = TRUE)
                         #& StoufferZ >= quantile((tConsensus_s$StoufferZ), 0.05, na.rm = TRUE)
                         ]

subset_dt <- subset_dt %>%
  mutate(Consistency_Range = cut(Consistency, 
                                 breaks = seq(0.5, 1.0, by = 0.5/7), 
                                 include.lowest = TRUE))
consistency_bins <- length(unique(subset_dt$Consistency_Range))

ggside(subset_dt, mapping = aes(x = StoufferZ/norm_constant, y = AvgRho, colour = Consistency, alpha = n_labels, size = n_labels)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_point() +
  scale_color_viridis_c(option = "turbo") +
  scale_size_continuous(range = c(0.2, 2.5)) + 
  scale_alpha_continuous(range = c(0.3, 0.8)) +
  #scale_color_gradient(low = "grey", high = "darkgreen") +
  labs(x = "Q0.95-Normalized Trans-Z", y = "Average Trans-Correlation") +
  
  anno_right(size = 0.3) +
  ggalign() +
  geom_density(aes(x = after_stat(density), y = AvgRho, 
                   colour = Consistency_Range), 
               position = "identity") +
  scale_color_manual(values = turbo(consistency_bins), guide = "none") +
  
  anno_top(size = 0.3) +
  ggalign() +
  geom_density(aes(x = StoufferZ/norm_constant, y = after_stat(density), 
                   colour = Consistency_Range), 
               position = "identity") +
  scale_color_manual(values = turbo(consistency_bins), guide = "none")
```

```{r}
#| label: get_type_C
norm_factor <- quantile(abs(tConsensus_t$StoufferZ), 0.95, na.rm = TRUE)
tConsensus_t[, Z_norm := StoufferZ / norm_factor]

C_t_tmp <- dcast(tConsensus_t, Ligand ~ Gene, value.var = "Z_norm", fill = 0)

C_t_mtx <- as.matrix(C_t_tmp[, -1])
rownames(C_t_mtx) <- C_t_tmp$Ligand

rm(C_t_tmp)
```

## 5.3 Omega tensor

```{r}
#| label: get_Omega_s

tryCatch({
  Omega_s <- build_Omega(phi = myPrism@phi_cellState@phi, 
                         ligand_target_matrix = (ligand_target_matrix), 
                         C_input = ((C_s_mtx)), 
                         lr_network = lr_network, 
                         mask_threshold = 0.5, 
                         masking_type = "soft")
  saveRDS(Omega_s, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "Omega_s.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved Omega_s successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting Omega_s failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```

```{r}
omega_s_df <- as.data.frame.table(Omega_s)
colnames(omega_s_df) <- c("Sender", "Receiver", "Ligand", "Potential")

omega_s_filtered <- omega_s_df %>% 
  filter(
    Potential != 0,
    Potential > quantile(Potential, 0.05, na.rm = TRUE),
    #Potential > -1,
    Potential < quantile(Potential, 0.95, na.rm = TRUE)
    #Potential < 1
  ) %>% 
  mutate(Receiver = as.character(Receiver)) %>% 
  arrange(Receiver) %>% 
  mutate(Receiver = factor(Receiver, levels = unique(Receiver)))
```

```{r}
ggplot(omega_s_filtered, aes(x = Potential, fill = Receiver, color = Receiver)) +
  geom_density(alpha = 0.5) +
  facet_grid(Receiver ~ ., scales = "free_y", switch = "y") +
  #geom_vline(xintercept = mean(omega_df_filtered$Potential), linetype = "dashed", color = "red") +
  theme_minimal() +
  scale_fill_manual(values = state_colors) +
  scale_color_manual(values = state_colors) +
  labs(title = "Distribution of Signed Interaction Potentials (Omega)",
       subtitle = "Filtered for non-zero interactions",
       x = "Regulatory Potential",
       y = "Count of Gene-State Pairs") +
  theme(
    strip.text.y.left = element_text(angle = 0, hjust = 1, size = 8), # Horizontal labels
    strip.background = element_blank(),
    panel.spacing = unit(0.1, "lines"), # Bring them close together
    axis.text.y = element_blank(),      # Hide the internal density axis
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

```{r}
#| label: get_Omega_t

tryCatch({
  Omega_t <- build_Omega(phi = myPrism@phi_cellType@phi, 
                         ligand_target_matrix = (ligand_target_matrix), 
                         C_input = ((C_t_mtx)), 
                         lr_network = lr_network, 
                         mask_threshold = 0.5, 
                         masking_type = "soft")
  saveRDS(Omega_t, file=paste0(data_dir, "NotebookRData/HuukiResiduals/", "Omega_t.rds"))
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = "Saved Omega_t successfully ✅",
       add_headers(Title = "R Process Success")
       )
}, error = function(e) {
  POST(paste0("https://ntfy.sh/", ntfy_topic),
       body = paste0("Getting Omega_t failed: ", e$message),
       add_headers(Title = "R Process ERROR ❌", Priotity = "high")
       )
})
```

```{r}
omega_t_df <- as.data.frame.table(Omega_t)
colnames(omega_t_df) <- c("Sender", "Receiver", "Ligand", "Potential")

omega_t_filtered <- omega_t_df %>% 
  filter(
    Potential != 0,
    Potential > quantile(Potential, 0.05, na.rm = TRUE),
    #Potential > -1,
    Potential < quantile(Potential, 0.95, na.rm = TRUE)
    #Potential < 1
  ) %>% 
  mutate(Receiver = as.character(Receiver)) %>% 
  arrange(Receiver) %>% 
  mutate(Receiver = factor(Receiver, levels = unique(Receiver)))
```

```{r}
ggplot(omega_t_filtered, aes(x = Potential, fill = Receiver, color = Receiver)) +
  geom_density(alpha = 0.5) +
  facet_grid(Receiver ~ ., scales = "free_y", switch = "y") +
  #geom_vline(xintercept = mean(omega_df_filtered$Potential), linetype = "dashed", color = "red") +
  theme_minimal() +
  scale_fill_manual(values = type_colors) +
  scale_color_manual(values = type_colors) +
  labs(title = "Distribution of Signed Interaction Potentials (Omega)",
       subtitle = "Filtered for non-zero interactions",
       x = "Regulatory Potential",
       y = "Count of Gene-Type Pairs") +
  theme(
    strip.text.y.left = element_text(angle = 0, hjust = 1, size = 8), # Horizontal labels
    strip.background = element_blank(),
    panel.spacing = unit(0.1, "lines"), # Bring them close together
    axis.text.y = element_blank(),      # Hide the internal density axis
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

```{r}
#| label: omega_summaries
niche_strength <- apply(Omega_s, c(1,2), sum)
gene_pressure <- apply(Omega_s, c(2,3), sum)
```

Visualization

```{r}
#| label: fig-niche_strength
#| fig-cap: "Heatmap of interaction strength across cell states."
ggheatmap(niche_strength) + 
  layout_title("Global Niche Interaction Strength") +
  labs(x = "Receiver", y = "Sender") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  anno_right(size = 0.2) +
  align_dendro()+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  anno_top(size = 0.3) +
  align_dendro()
```

## 5.4 Weighted Omega Tensor

```{r}
#| label: weigh_omega
avg_props <- table(sce$cellType_hc) / ncol(sce)
Weighted_Omega <- Omega_s

for(k in all_states){
  Weighted_Omega[k, , ] <- Weighted_Omega[k, , ] * avg_props[k]
}

niche_strength_weighted <- apply(Weighted_Omega, c(1,2), sum)
weighted_gene_pressure <- apply(Weighted_Omega, c(2,3), sum)
```

Visualization

```{r}
#| label: fig-weighted_niche_strength
#| fig-cap: "Heatmap of weighted interaction strength across cell states."
ggheatmap(niche_strength_weighted) + 
  layout_title("Effective Niche Pressure") +
  labs(x = "Receiver", y = "Sender") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  anno_right(size = 0.2) +
  align_dendro() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  anno_top(size = 0.3) +
  align_dendro()
```

# 6. Checking gene pressure effect on BayesPrism residuals

Making a dataframe containing residual and gene pressure data for all states

```{r}
#| label: residual_test_df
results_list <- list()
common_genes <- intersect(colnames(gene_pressure), colnames(residual_mtx))

for(type in names(bp_res@prism@map)){
  for (state in bp_res@prism@map[[type]]){
    
    # Check if state exists in the matrix and has non-NA data
    if (state %in% rownames(residual_mtx) && sum(!is.na(residual_mtx[state, ])) != 0) {
      # Filter to expressed genes per cell type
      expressed_genes <- intersect(common_genes, colnames(true_cpm_mtx)[true_cpm_mtx[state, ] >= quantile(true_cpm_mtx[state,], 0.1)])
      
      tmp <- data.frame(
        type = type,
        state = state,
        gene = expressed_genes,
        gene_pressure = as.numeric(gene_pressure[state, expressed_genes]),
        weighted_gene_pressure = as.numeric(weighted_gene_pressure[state, expressed_genes]),
        residual = as.numeric(residual_mtx[state, expressed_genes]),
        stringsAsFactors = FALSE
      )
      
      results_list[[length(results_list) + 1]] <- tmp
    }
  }
}

aggregated_df <- do.call(rbind, results_list)
```

Making a dataframe for plotting

```{r}
#| label: plot_df
types_to_plot <- c(
  "Astro", 
  "EndoMural", 
  "Micro", 
  "Oligo", 
  "OPC", 
  "Inhib",
  "Excit"
  )
#types_to_plot <- c("OPC")
#states_to_plot <- c("Excit_09", "Excit_05", "Excit_10")

plot_df <- aggregated_df[aggregated_df$type %in% types_to_plot 
                         #& aggregated_df$residual != 0
                         , ]
#plot_df$residual <- abs(plot_df$residual)
#plot_df$gene_pressure <- abs(plot_df$gene_pressure)
#plot_df$weighted_gene_pressure <- abs(plot_df$weighted_gene_pressure)

plot_df <- plot_df[(plot_df$gene_pressure) < quantile(plot_df$gene_pressure, 0.99)
                   & (plot_df$gene_pressure) > quantile(plot_df$gene_pressure, 0.01)
                   #& plot_df$residual < quantile(plot_df$residual, 0.95)
                   #& plot_df$residual > quantile(plot_df$residual, 0.05)
                   , ]

# test
plot_df$test <- sample(plot_df$gene_pressure)

# DIRECTIONALITY
#plot_df$residual <- abs(plot_df$residual)
#plot_df$gene_pressure <- abs(plot_df$gene_pressure)
#plot_df$weighted_gene_pressure <- abs(plot_df$weighted_gene_pressure)
#directional_mask <- plot_df$residual/abs(plot_df$residual)
#plot_df$gene_pressure <- plot_df$gene_pressure * directional_mask
#plot_df$weighted_gene_pressure <- plot_df$weighted_gene_pressure * directional_mask
#lot_df$test <- plot_df$test * directional_mask

cor.test((plot_df$residual), plot_df$gene_pressure, method="spearman")
```

Gene pressure plots

```{r}
#| label: fig-gene_pressure
#| fig-cap: !expr paste0("Effect of cell-state-specific gene pressure on residuals between the BayesPrism-generated gene expressions and the observed snRNA-seq expression in sample ", sample, ". The top plot top shows the mean errors binned by gene pressure score ranges. Notably, a positive trend can be observed, which is supported by the spearman correlation rho = ", cor((plot_df$residual), plot_df$gene_pressure, method="spearman"), ". The bottom plot shows the scatter plot for gene pressure vs. error.")

ggside(plot_df, mapping = aes(gene_pressure, (residual), colour = type)) -
  scheme_theme(theme_bw()) +
  layout_theme(panel.border = element_blank()) +
  geom_point(size = 1) +
  scale_color_manual(values = type_colors) +
  geom_smooth(method = "lm", color = "darkgray") +
  labs(y = "Error") + 
  # Density plot
  anno_bottom(size = 0.3) +
  ggalign() +
  geom_density(aes(x = gene_pressure, y = after_stat(density), colour = type), position = "identity") +
  scale_color_manual(values = type_colors, guide = "none") +
  #labs(title = paste("Effect of Gene Pressure on BayesPrism Error"))+
  # Density plot
  anno_right(size = 0.1) +
  ggalign() +
  geom_density(aes(x = after_stat(density), y = residual, colour = type), position = "identity") +
  scale_color_manual(values = type_colors, guide = "none") +
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5, hjust = 0)) +
  # Bar plot
  anno_top(size = 1) +
  ggalign() +
  stat_summary_bin(
    mapping = aes(x = gene_pressure, y = (residual), fill = after_stat(count)),
    bins = 50,
    geom = "bar",
    fun.data = function(x) {
      data.frame(
        y = mean(x),
        count = length(x)
      )
    }
  ) +
  scale_fill_gradient(low = "lightgrey", high = "#08306b", name = "Gene Count") +
  labs(x = "Gene Pressure", y = "Mean Error")
```

Weighted gene pressure plots

```{r}
#| label: fig-weighted_gene_pressure
#| fig-cap: !expr paste0("Effect of weighted cell-state-specific gene pressure on residuals between the BayesPrism-generated gene expressions and the observed snRNA-seq expression in sample ", sample, ". rho = ", cor((plot_df$residual), plot_df$weighted_gene_pressure, method="spearman"), ".")

cor.test((plot_df$residual), plot_df$weighted_gene_pressure, method="spearman")

ggside(plot_df, mapping = aes(weighted_gene_pressure, (residual), colour = type)) -
  scheme_theme(theme_bw()) +
  layout_theme(panel.border = element_blank()) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_color_manual(values = type_colors) +
  geom_smooth(method = "lm", color = "darkgray") +
  labs(x = "Weighted Gene Pressure", y = "Error") +
  # Bar plot
  anno_top(size = 1) +
  ggalign() +
  stat_summary_bin(
    mapping = aes(x = weighted_gene_pressure, y = (residual), fill = after_stat(count)),
    bins = 50,
    geom = "bar",
    fun.data = function(x) {
      data.frame(
        y = mean(x),
        count = length(x)
      )
    }
  ) +
  scale_fill_gradient(low = "lightgrey", high = "#08306b", name = "Gene Count") +
  labs(title = paste("Effect of Weighted Gene Pressure on BayesPrism Error"), 
       y = "Mean Error")
```
